Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-07-26 paper.tex
Add HMNS test text

The volume data are components of the metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
and the hydrodynamical data $W$, $u_i$, $\varrho$, $T$, and $Y_e$.
These fields are stored in spectral representation.
If they are computed via hydrodynamical simulation, they are saved to disk
at specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a quasi-stationary solution to the
\todo{clarify quasistationary timescale}
general relativistic hydrodynamics equations,
a single time is stored to disk, and no time interpolation is needed.

@@

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).
The opacity table is computed using \lstinline{NuLib} from cross-sections
neglecting weak magnetism
\todo{anything else?}
and is stored on a grid covering energy, density, temperature, and composition
ranges spanning $\varepsilon_i\in[1,280.5]$~MeV logarithmically,
$\varrho\in[10^6,6.3\times10^15]\,{\rm g}\,{\rm cm}^{-3}$ logarithmically,
$T\in[0.05,200]$~MeV logarithmically, and 
$Y_e\in[0.035,0.55]$ linearly,
with grid extents$\{18,82,65,51\}$ respectively.
Interpolation is performed linearly.

@@

Such a configuration was modeled in \cite{fouc2016-m1_nsns} by
evolving the final inspiral and merger of two identical neutron stars of
isolated gravitational mass 1.2~$M_{\odot}$. The postmerger configuration
was simulated using a gray M1 transport scheme for the neutrinos, evolving the
energy density, number density, and energy flux \cite{fouc2016-m1_evolve_n}
for $\sim10\,{\rm ms}$ following merger.
The fluid was modeled using the LS220 equation of state \cite{latt1991-nuc_eos}.

We use a single time snapshot of the volume data from that configuration
at $t=11\,{\rm ms}$ after merger (see
Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat})
sampling the neutrino fields with ray tracing to compare to some
measures of the field computed from the M1 transport simulation.
\todo{here focus on plots as functions of $\theta$}

@@

In Tab.~\ref{tab:nsns_rt_vs_m1} we present total and relative number and
energy luminosities for this configuration.
With scattering turned off, ray tracing errors are as large as 250\%
(for heavy lepton energy luminosities).
But with scattering turned on, agreement is better than 20\%, and
in most cases much better.
\todo{get accurage M1 numbers from Francois}

\begin{table}%[h]
  \caption{
    Comparison of total luminosities and average energies between
    ray tracing using the \emph{noscat} method `RT noscat',
    ray tracing using the \emph{gray} \emph{scat} method `RT scat',
    and the M1 transport simulation `M1'.
    The ray tracing totals were computed from sums over
    observers placed in the $y$-$z$ plane along an arc at coordinate distance
    $r=250\,{\rm km}$ from the center of the star, at 30 positions
    distributed evenly in $\cos\theta$ over the colatitude range
    $\theta\in[0,90^{\circ}]$.
    M1 values are taken from \cite[Figs. 7, 9, 10]{fouc2016-m1_evolve_n}.
    Energy luminosities have units $10^{52}\,{\rm erg}\,{\rm s}^{-1}$,
    number luminosities $10^{57}\,{\rm s}^{-1}$, and
    average energies MeV.
    The $\nu_x$ luminosities are per-species:
    multiply by four to get the total heavy-lepton neutrino luminosities.
  }
  \label{tab:nsns_rt_vs_m1}
  \begin{tabularx}{\columnwidth}{X r r r}
    & {\bf RT noscat} & \,\,{\bf RT scat} & \,\,{\bf M1} \\
    \hline
    $L_{\nu_e}$                                 & 5.95 & 6.01 & 5 \\
    $L_{\bar{\nu}_e}$                           & 10.6 & 11.8 & 12 \\
    $L_{\nu_x}$                                 & 30.2 & 12.6 & 12 \\
    $R_{\nu_e}-R_{\bar{\nu}_e}$                 & -1.06 & -1.86 & -2 \\
    $\langle \varepsilon_{\nu_e} \rangle$       & 12.08 & 11.88 & 12 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle$ & 15.96 & 14.69 & 15 \\
    $\langle \varepsilon_{\nu_x} \rangle$       & 36.25 & 23.67 & 26 \\
    \hline
  \end{tabularx}
\end{table}

>>>>

The volume data are components of the metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
\todo{state num independent components}
and the hydrodynamical data $W$, $u_i$, $\varrho$, $T$, and $Y_e$.
These fields are stored in spectral representation.
If they are computed via hydrodynamical simulation, they are saved to disk
at specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a quasi-stationary solution to the
\todo{clarify quasistationary timescale}
general relativistic hydrodynamics equations,
a single time is stored to disk, and no time interpolation is needed.

@@

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).
The opacity table is computed using \lstinline{NuLib} from cross-sections
neglecting weak magnetism
\todo{anything else?}
and is stored on a grid covering energy, density, temperature, and composition
ranges spanning $\varepsilon_i\in[1,280.5]$~MeV logarithmically,
$\varrho\in[10^6,6.3\times10^{15}]\,{\rm g}\,{\rm cm}^{-3}$ logarithmically,
$T\in[0.05,200]$~MeV logarithmically, and 
$Y_e\in[0.035,0.55]$ linearly,
with grid extents$\{18,82,65,51\}$ respectively.
Interpolation is performed linearly.

@@

Such a configuration was modeled in \cite{fouc2016-m1_nsns} by
evolving fluid and spacetime through the
final inspiral and merger of two identical neutron stars of
isolated gravitational mass 1.2~$M_{\odot}$. The postmerger configuration
was simulated using a gray M1 transport scheme for the neutrinos, evolving the
energy density, number density, and energy flux \cite{fouc2016-m1_evolve_n}
for $\sim10\,{\rm ms}$ following merger.
The fluid was modeled using the LS220 equation of state \cite{latt1991-nuc_eos}.

We use a single time snapshot from that configuration
at $t=11\,{\rm ms}$ after merger.
The volume data are represented on a pseudospectral domain of a
sphere with concentric shells extending to $r\approx1400$~km.
Radial grid spacings are $\Delta r \approx 0.15$~km in the star
and $\Delta r \approx 2.5$~km in the disk,
with 12 cells spanning polar angles $\theta\in[0,\pi]$
and 24 cells spanning azimuthal angles $\phi\in[0,2\pi]$.
\todo{way too-low angular resolution}
Though a pseudospectral representation of non-smooth hydrodynamic data
introduces some Gibbs-like oscillations in the variables,
we choose to use this representation instead of the mesh-refined
finite difference grid of the evolution because the pseudospectral
representation uses much less memory:
95~MB in pseuspectral vs.\ 3~GB in finite difference representation.
Obviously, memory loads of this order are not insurmountable;
but they would require some modifications to our
volume data interpolation infrastructure.
For the purposes of this analysis the pseudospectral representation is adequate.
Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat}
show slices of density and temperature from the finite difference volume data.

@@

Opacities are computed from the LS220 equation of state using
\lstinline{NuLib} and are stored as a table covering
energy, density, temperature, and composition ranges
spanning $\varepsilon\in[1,280.5]$~MeV logarithmically,
$\varrho\in[10^6,3.2\times10^{15}]\,{\rm g}\,{\rm cm}^{-3}$ logarithmically,
$T\in[0.05,150]$~MeV logarithmically, and 
$Y_e\in[0.035,0.55]$ linearly,
with grid extents$\{18,82,65,51\}$ respectively.

We place stationary (i.e.\ Eulerian) observers at fixed coordinate radius
$r=250$~km in the $y$-$z$ plane along an arc at $N_\theta$ positions
distributed linearly in $\cos\theta$ over the northern hemisphere
using $\theta_i=\cos^{-1}\left(1-i/(N_\theta-1)\right)$
for $i=0,1,2,\ldots N_\theta-1$.
We choose $N_\theta=30$.

Though the ray tracing sampling of $f$ is done in full general relativity,
in our calculation of moments at the observer positions,
we make the simplifying assumption of Minkowski spacetime.
This assumption introduces errors in our moments of order
$GM/r=4\,{\rm km}/250\,{\rm km}$
\todo{too simplistic}
(with $G$ the gravitational constant and $M$ the central mass),
or $\sim 1\%$.

Each observer samples the distribution function over a uniform grid
in energy, cosine of polar angle, and azimuthal angle with extents
$\{N_{\tilde{\varepsilon}},N_A, N_B\}$ spanning the ranges
$\tilde{\varepsilon}\in(0,100)$~MeV,
$\cos A\in\left((\cos A)_{\rm min}, 1\right)$, and
$B\in[0,2\pi)$.
Using the midpoint rule,
the radial fluxes of number and energy from Eqns.~\ref{eqn:Ha}
and \ref{eqn:J_H_S_eps_integrated} become sums:
\begin{align}
  \label{eqn:Kr_minkowski}
  \tilde{K}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^2 f_m,\\
  \label{eqn:Hr_minkowski}
  \tilde{H}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^3 f_m,
\end{align}
with $f_m \equiv f(\tilde{\varepsilon}_m,\cos A_m, B_m)$,
$\Delta \equiv \Delta\tilde{\varepsilon}\, \Delta(\cos A)\, \Delta B$, and
$N \equiv N_{\tilde{\varepsilon}}\, N_A\, N_B$.
Average energies are given by
$\langle\tilde{\varepsilon}\rangle(\theta)=H^r(\theta)/K^r(\theta)$.
We choose $N_{\tilde{\varepsilon}}=30$, $N_A=120$, and $N_B=20$.
For the \emph{scat} case we choose $(\cos A)_{\rm min}=0.81$;
For the \emph{noscat} case, presenting a smaller emitting region,
we choose $(\cos A)_{\rm min}=0.91$.

We also combine measurements from all observers to estimate
total luminosities and averages over the sky.
We assume the disk is azimuthally symmetric and
symmetric with respect to reflection across the equatorial plane.
Figs.~\ref{fig:nsns_rho_merid} and \ref{fig:nsns_rho_equat} show how these
assumptions are approximately valid at 11~ms after merger.
Total luminosities are then computed as integrals of the radial fluxes
over $\cos\theta$.
Using the trapezoid rule the number and energy luminosities become sums:
\begin{align}
  \label{eqn:luminosity_R}
  R &= C
  \left(\frac{1}{2}(\tilde{K}^r_0+\tilde{K}^r_{N_\theta-1})
  +\sum\limits_{m=1}^{N_\theta-2}\tilde{K}^r_m\right),\\
  \label{eqn:luminosity_L}
  L &= C
  \left(\frac{1}{2}(\tilde{H}^r_0+\tilde{H}^r_{N_\theta-1})
  +\sum\limits_{m=1}^{N_\theta-2}\tilde{H}^r_m\right),
\end{align}
with $C \equiv 4\pi r^2\, \Delta(\cos \theta)$,
\todo{source of GR error in coordinate $r$}
$\tilde{K}^r_m \equiv \tilde{K}^r(\theta_m)$, and
$\tilde{H}^r_m \equiv \tilde{H}^r(\theta_m)$.
Average energies over the whole sky are then
$\langle\tilde{\varepsilon}\rangle=L/R$.

\begin{table}%[h]
  \caption{
    Comparison of total luminosities and average energies
    of the hypermassive neutron star configuration.
    The methods are ray tracing using the \emph{noscat} method `RT noscat',
    ray tracing using the \emph{gray} \emph{scat} method `RT scat',
    and the M1 transport simulation `M1'.
    The ray tracing totals were computed from sums over
    observers placed in the $y$-$z$ plane.
    M1 values are taken from \cite[Figs. 7, 9, 10]{fouc2016-m1_evolve_n}.
    Energy luminosities have units $10^{52}\,{\rm erg}\,{\rm s}^{-1}$,
    number luminosities $10^{57}\,{\rm s}^{-1}$, and
    average energies MeV.
    The $\nu_x$ luminosities are per-species:
    multiply by four to get the total heavy-lepton neutrino luminosities.
  }
  \label{tab:nsns_rt_vs_m1}
  \begin{tabularx}{\columnwidth}{X r r r}
    & {\bf RT noscat} & \,\,{\bf RT scat} & \,\,{\bf M1} \\
    \hline
    $L_{\nu_e}$                                 & 5.95 & 6.01 & 5 \\
    $L_{\bar{\nu}_e}$                           & 10.6 & 11.8 & 12 \\
    $L_{\nu_x}$                                 & 30.2 & 12.6 & 12 \\
    $R_{\nu_e}-R_{\bar{\nu}_e}$                 & -1.06 & -1.86 & -2 \\
    $\langle \varepsilon_{\nu_e} \rangle$       & 12.08 & 11.88 & 12 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle$ & 15.96 & 14.69 & 15 \\
    $\langle \varepsilon_{\nu_x} \rangle$       & 36.25 & 23.67 & 26 \\
    \hline
  \end{tabularx}
\end{table}

Tab.~\ref{tab:nsns_rt_vs_m1} compares the all-sky
\todo{get accurate M1 numbers from Francois}
luminosities and average energies from ray tracing and from
the M1 transport simulation,
which serves as our fiducial calculation.
Because the model was evolved with a gray M1 transport scheme,
the only \emph{scat} method we can use in our ray tracing is the
\emph{gray} method.
As in the post-bounce configuration in Sec.~\ref{ssec:test_collapse},
the \emph{scat} treatment is more faithful than the \emph{noscat} treatment,
significantly so for the heavy-lepton neutrinos which are most affected.
With scattering turned off ray tracing overestimates $L_{\nu_x}$ by 250\%;
but with scattering turned on agreement in all measures
is better than 20\%, and in most cases much better.
Note that $R_{\bar{\nu}_e}$ dominates over $R_{\nu_e}$
\todo{Need $R$ magnitude not just difference}
because the disrupted neutron star material is releptonizing

The distributions of radial fluxes over observer position are shown
in Figs.~\ref{fig:nsns_theta_distrib_R} and \ref{fig:nsns_theta_distrib_L}.
With the \emph{noscat} treatment the $\nu_x$ luminosities are relatively
constant in $\theta$, because the disk is optically thin to $\nu_x$,
and most of the neutrinos come from the hypermassive neutron star,
which is roughly spherical.
The slight upward trend in $\nu_x$ luminosities with $\theta$ may be due to
asymmetries in the configuration.
It is not due to the Doppler shift from the rapid rotation of the star
(a hypothesis we tested by setting $u_i=0$ and $W=1$ in the ray
tracing equations).
Because of their larger absorption cross section,
the disk is not optically thin to $\nu_e$ and $\bar{\nu}_e$,
and observers within view of the hot hypermassive neutron star measure
the largest radial fluxes.

When we turn on scattering, the disk is no longer transparent to $\nu_x$,
and $\nu_x$ luminosities present the same qualitative $\theta$-dependence
as that of the other species.
In general, scattering increases luminosities for polar observers
($\theta \sim 0^\circ$) because additional high energy neutrinos
from the hypermassive neutron star are scattered by the disk back up to
the observer.
For observers in the equatorial plane ($\theta \sim 90^\circ$)
the effect of scattering is a decrease in energy luminosities,
since some of the highest energy neutrinos from deep in the interior
are scattered away from the line of sight by the disk.

Fig.~\ref{fig:nsns_theta_distrib_avg_eps} shows the distribution of
average energies of radial fluxes over observer positions.
With and without scattering, the average energies are highest
for observers near the polar axis, since polar observers get a
direct view of the hot hypermassive neutron star.
Scattering decreases average energies across all observer positions,
as it did in the post-bounce configuration, or any configuration of
a hot interior surrounded by a scattering envelope.
The strength of the effect of scattering on the average energies of
the different species is seen to follow the ranking
$\nu_x>\bar{\nu}_e>\nu_e$ due to two factors:
the average energies of the spectra obey the same ranking,
and the thicknesses of the different species' scattering envelopes
also obey the same ranking.
As we showed in the optically thin limit of
Eqn.~\ref{eqn:fse_homogeneous_thin_limit},
\todo{too simplistic}
the scattering contribution $f_{\rm SE}$ is
proportional to energy $\varepsilon^2$ and path length $s$.
